---
title: "STAT 306 Project"
format: html
editor: visual
---

## Introduction

The analysis utilizes the **GBSG breast cancer dataset** in Royston and Altman (2013) to analyze the relationship between progresterone receptor levels (pgr), patient age, and tumor size at the time of surgery. The document is organized into the following:

-   Data Cleaning

-   Exploratory Data Analysis (EDA)

-   Linear Regression Analysis

-   Hypothesis Testing

-   Principal component analysis

## 1. Data Cleaning

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(GGally)
```

```{r}
# Load dataset
breast_cancer <- read.csv("data/gbsg.csv")
head(breast_cancer)
```

```{r}
# Find the data dimension
dim(breast_cancer)
```

The data consists of 686 observations.

```{r}
# Verify patient id uniqueness
pid <- breast_cancer |>
  summarize(total = length(unique(pid)))
pid
```

```{r}
# Remove the row index column (first column) and non-baseline variables
breast_cancer <- breast_cancer[,-1] |>
  select(-hormon,-rfstime,-status, -pid)
head(breast_cancer)
```

```{r}
# Find the variable name & types
str(breast_cancer)
```

```{r}
# Adjust variable type for categorical variable
breast_cancer$meno <- as.factor(breast_cancer$meno)
breast_cancer$grade <- as.factor(breast_cancer$grade)
# Check the variable type
str(breast_cancer)
```

```{r}
# Find the number of missing values for each variable in the dataset
missing_val_count <- colSums(is.na(breast_cancer))
missing_val <- data.frame(missing_val_count) %>%
    mutate(percentage = round(missing_val_count/nrow(breast_cancer) * 100,2))
missing_val
```

```{r}
# Verify class imbalances for categorical variables
meno <- breast_cancer |>
  group_by(meno) |>
  summarize(total = n())
meno
grade <- breast_cancer |>
  group_by(grade) |>
  summarize(total = n())
grade
```

## 2. Exploratory Data Analysis (EDA)

```{r}
pairs(breast_cancer)
ggpairs(breast_cancer, axisLabels = "none", progress = FALSE)
```

```{r}
boxplot(size ~ meno, data = breast_cancer) +
  labs(title = "Distribution of tumor size based on menopausal status", 
       x = "menopausal status",
       y = "tumor size (mm)")
boxplot(size ~ grade, data = breast_cancer)
plot(size ~ pgr, data = breast_cancer)
plot(size ~ er, data = breast_cancer)
plot(size ~ age, data = breast_cancer)
plot(size ~ log(nodes), data = breast_cancer)

# Show no linear relationship
plot(size ~ age, data = breast_cancer)
# Fix with log(Y)
plot(log(size) ~ age, data = breast_cancer)

plot(size ~ pgr, data = breast_cancer)
plot(log(size) ~ log(pgr), data = breast_cancer)

```

## 3. Linear Regression Analysis

```{r}
full_model <- lm(size ~ ., breast_cancer)
summary(full_model)
model1 <- lm(size ~ age * pgr, breast_cancer)
summary(model1)
model2 <- lm(size ~ age, breast_cancer)
summary(model2)

```

\

\
