---
title: "STAT 306 Project"
format: html
editor: visual
---

## Introduction

The analysis utilizes the **GBSG breast cancer dataset** in Royston and Altman (2013) to analyze the relationship between progresterone receptor levels (pgr), patient age, and tumor size at the time of surgery. The document is organized into the following:

-   Data Cleaning

-   Exploratory Data Analysis (EDA)

-   Linear Regression Analysis

-   Hypothesis Testing

-   Principal component analysis

## 1. Data Cleaning

Although the given data is complete (no missing value), some adjustments are .

```{r}
# Load required libraries
library(dplyr)
```

```{r}
# Load dataset
breast_cancer <- read.csv("data/gbsg.csv")
head(breast_cancer)
```

```{r}
# Remove the row index column (first column)
breast_cancer <- breast_cancer[,-1]
head(breast_cancer)
```

```{r}
# Find the data dimension
dim(breast_cancer)
```

```{r}
# Find variable name & type
str(breast_cancer)
```

Since all the variables are integer and some of them appears to be a binary variable, let's switch it to a factor type.

```{r}
# Adjust variable type for categorical variable
breast_cancer$meno <- as.factor(breast_cancer$meno)
breast_cancer$grade <- as.factor(breast_cancer$grade)
breast_cancer$hormon <- as.factor(breast_cancer$hormon)
breast_cancer$status <- as.factor(breast_cancer$status)
# Check the variable type
str(breast_cancer)
```

```{r}
# Find the number of missing values for each variable in the dataset
missing_val_count <- colSums(is.na(breast_cancer))
missing_val <- data.frame(missing_val_count) %>%
    mutate(percentage = round(missing_val_count/nrow(breast_cancer) * 100,2))
missing_val
```

```{r}
# Find class imbalance for categorical variable
meno <- breast_cancer |>
  group_by(meno) |>
  summarize(total = n())
meno
grade <- breast_cancer |>
  group_by(grade) |>
  summarize(total = n())
grade
hormon <- breast_cancer |>
  group_by(hormon) |>
  summarize(total = n())
hormon
status <- breast_cancer |>
  group_by(status) |>
  summarize(total = n())
status
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

## 3. Linear Regression Analysis

```{r}

```
